Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
\[(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi\]
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
\jb{do we? I'd think that the admissibility criterion below is enough.}
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{conjecture} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
it would terminate when seen as a functional program for \relax{some}
evaluation order. For example,
$\{\forallf{\con{f}} n : {\typeint}.\;\, \lite\bigl(
n \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)\}$
can be shown well-founded under an evaluation strategy that consides one of the
two $\lite$ branches, based on the result of evaluating the condition.
Logically, such dependencies can be captured by congruence rules. Krauss
develops these ideas systematically in the more general context of higher-order
logic \cite[Section 2]{krauss-2009-phd}.

Another interesting fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines. The
following theorem connects these notion with admissibility.

\begin{theorem}
Let
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n}
\vec x.\; \varphi_n \}
$ be a set of function definitions such that for every
$(\stypes{\Sigma},\, \sfuns{\Sigma} - \sfundefs{\Sigma})$-interpretations
$\I_0$, there exists a unique
$\Sigma$-interpretation $\I$ that extends $\I_0$
and for which $\I \models \DDD$.
Then the set\/ $\DDD$ is admissible.
\end{theorem}
\begin{proofsketch}
Let $\varphi$ be a $T$-satisfiable formula closed under function expansion with
respect to $\DDD$.

@@

The higher-order model finder Nitpick for the Isabelle/HOL proof assistant uses
another variant of Kuncak and Jackson's approach inside a Kleene-style
three-valued logic, inspired by abstract interpretation. It was also the first
tool of its kind to corecursion and codatatypes
\cite{blanchette-2013-relational}. The three-valued logic approach extends each
approximated type with an unknown value, which is propagated by function
application. This approach works reasonably well in Nitpick, because it builds
on Kodkod's relational logic, but our initial experiments with \cvc convinced
us that it is more efficient to avoid unknown values by adding constraints.

@@

In future work, it would be interesting to evaluate the approach against other
model finders, notably Leon \ct{?? Leon is a model finder? Alloy?} and Nitpick, and enrich the benchmark suite with
problems exercising \cvc's support for codatatypes
\cite{reynolds-blanchette-2015-codata}. We also plan to integrate \cvc as a
counterexample generator in proof assistants.
\ct{Further work could also include identifying simple and/or syntactic sufficient conditions for admissibility. }

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
We would like to thank Damien Busato-Gaston and Emmanouil Koukoutos for
providing the initial set of benchmarks used in the evaluation.
}

>>>>

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
\[(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi\]
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
\jb{do we? I thought that the admissibility criterion below is enough.}
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{conjecture} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
it would terminate when seen as a functional program for \relax{some}
evaluation order. For example, the definition~(\ref{eq:pii}) from the
introduction can be shown well-founded under an evaluation strategy that
consides one of the two $\lite$ branches, based on the result of evaluating the
condition. Logically, such dependencies can be captured by congruence rules.
Krauss develops these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

Another useful fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines. The
following theorem connects totality with admissibility.

\begin{definition}\rm
Let $\DDD$ be a set of function definitions for $\sfundefs{\Sigma}$. The set
$\DDD$ is \emph{total} if for every
$(\stypes{\Sigma},\, \sfuns{\Sigma} - \sfundefs{\Sigma})$-interpretations
$\I_0$, there exists a unique
$\Sigma$-interpretation $\I$ that extends $\I_0$
and for which $\I \models \DDD$.
\end{definition}

\begin{theorem}
If
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n}
\vec x.\; \varphi_n \}
$ is a total set of function definitions, then it is admissible.
\end{theorem}
\begin{proofsketch}
Let $\varphi$ be a $T$-satisfiable formula closed under function expansion with
respect to $\DDD$.

@@

The higher-order model finder Nitpick for the Isabelle/HOL proof assistant uses
another variant of Kuncak and Jackson's approach inside a Kleene-style
three-valued logic, inspired by abstract interpretation. It was also the first
tool of its kind to corecursion and coalgebraic datatypes
\cite{blanchette-2013-relational}. The three-valued logic approach extends each
approximated type with an unknown value, which is propagated by function
application. This approach works reasonably well in Nitpick, because it builds
on Kodkod's relational logic, but our initial experiments with \cvc convinced
us that it is more efficient to avoid unknown values by adding constraints.

@@

In future work, it would be interesting to evaluate the approach against other
counterexample generators, notably Leon and Nitpick, and enrich the benchmark
suite with problems exercising \cvc's support for coalgebraic datatypes
\cite{reynolds-blanchette-2015-codata}. We also plan to integrate \cvc as a
counterexample generator in proof assistants. Finally, further work could also
include identifying further sufficient conditions for admissibility, thereby
enlarging the scope of the scheme presented here.

\ct{?? Leon is a model finder? Alloy?}
\jb{Alloy has no notion of datatypes, so it's pretty useless. Nitpick implements
datatypes and a more FOL-like logic (actually, HOL) on top of Alloy's backend,
using the Kuncak \& Jackson idiom for Alloy, so that's as close as we can get
to evaluating Alloy itself. Leon is, in part, a counterexample generator.
``Model finder'' is probably a stretch.}

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
Viktor Kuncak and Stephan Merz have made this work possible. We would also like
to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial
set of benchmarks used in the evaluation.
}
